---
title: "PCWG_share_01_main"
author: "Andy Clifton"
date: '`r Sys.Date()`'
output: pdf_document
---
<!-- ## Reading the source code?
This is R Markdown. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com> or <http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html>.

This script is designed to be used with RStudio. When you click the **Knit** button in RStudio a document will be generated that includes the output of any embedded R code chunks within the document. -->

# Introduction
This document demonstrates the concept of Literate Programming. An output PDF document and results are generated using the programing language `R` from the _main.rmd_ file and can be run by participants themselves.

## How to use main.rmd
install R (<http://www.r-project.org>) and Rstudio (<http://www.rstudio.com>), and then create a directory with all of the code and files (see below). When you click the **Knit** button in RStudio a document will be generated that includes text and results from the code embedded in _main.rmd_.

```{r clean up, echo = FALSE}
rm(list = ls())
```
## User Inputs
The _project.root_ variable defines the location of the files required for this analysis. The _made.by_ variable forms part of a label that will be added to the plots. 

The following user inputs were used in the preparation of this document:

```{r user-defined options}
# Where can files be found?
project.root <- file.path('/Users/aclifton/Documents/confidential/projects/PCWG Share 01')

# Who ran this script
made.by = "A. Clifton"
```
We now generate a directory (_output.dir_) to capture the outputs from this analysis.

```{r check old data actually exists, results="asis", echo = FALSE}
# define the path to the directory that we will use to store all of the data
output.dir = file.path(project.root, 
                       "analysis",
                       "all")
# create this directory if it doesn't already exist
dir.create(output.dir, 
           showWarnings = FALSE,
           recursive = TRUE)

```

## Packages
### Standard Packages
This script requires the _ggplot2_, _grid_, _knitr_, _RColorBrewer_, _rgdal_, _stringr_, and _XLConnect_ packages to run. These are called from the script but you may need to install them directly. For details of how to install packages, see the RStudio help. 

```{r load packages, message=FALSE, echo = FALSE}
require(ggplot2)

if(packageVersion("ggplot2") < "2.2.0") {
    stop("Need package:ggplot2 2.2 or higher for labs function")
}

require(grid)
require(XLConnect)
require(knitr)
require(reshape2)
require(stringr)
```

## Directory structure

```{r define file locations, message=FALSE, echo = FALSE}
# define the working directory
working.dir <- project.root
setwd(working.dir)

#identify data directory
data.dir = file.path(project.root,
                     "data")

# define where functions live
code.dir = file.path(project.root,
                     "code")

# source these functions
code.files = dir(code.dir, pattern = "\\.R$")
for (file in code.files){
  source(file = file.path(code.dir,file))
}

```
The folowing files should be placed in the _project.root_ directory:

* main.Rmd
* /__analysis__ directory containg results of the analysis
* /__code__ directory containing functions required for the analysis
* /__data__ directory containing the data files to be analyzed. 

```{r configure graphics, message=FALSE, echo = FALSE}
# configure graphics appearance
theme_set(theme_Literate(base_size = 8, 
               base_family = "sans"))
```

# Results from each data set
We now analyse the data from the simple data set. The plots are saved to the _analysis_ directory. If _data.public_ is FALSE, plots will be created for every data set. If _data.public_ is TRUE, only the final, aggregated data plots will be created.

```{r load individual data files, results='asis', echo = FALSE}
# set a counter to check for numbers of files that are data.version.earliest or later
count.version = 0
if (data.analyze.raw == TRUE){
  # identify the data sets that we have available
  data.files = dir(data.dir, 
                   recursive = TRUE,
                   pattern = "\\.csv$")
  
  for (data.file in data.files){
    # Read this data set;
  
    }
  }
}
```


``` {r save data, echo=FALSE}
# define the path to the directory that we will use to store all of the data
output.dir = file.path(project.root, 
                       "analysis",
                       "all")

# save the data
if (data.analyze.raw == TRUE){
  save(list = c("project.root",
                "made.by",
                "data.public",
                "data.version.earliest",
                "working.dir",
                "output.dir",
                "all.data"), 
       file = file.path(output.dir,"AggregatedData.RData"),
       envir = .GlobalEnv)
} else {
  load(file = file.path(output.dir,"AggregatedData.RData"))
}
```
